Optimal Multiway Spatial Joins

### 2024-05-16 msj.tex
yufei

\def\agm{\mathrm{AGM}}
\def\att{\mathbf{att}}
\def\actdom{\mathbf{actdom}}
\def\dom{\mathbf{dom}}
\def\gat{\mit{gat}}
%\def\id{\mathit{id}}
%\def\In{\mathrm{IN}}
\def\heavy{\mit{heavy}}
\def\join{\mathit{Join}}
\def\key{\mathit{key}}
%\def\light{\mathit{light}}
\def\light{\ell}
\def\load{\mathit{load}}
%\def\no{\mathit{no}}
\def\occ{\mathrm{OCC}}
\def\Out{\mathrm{OUT}}
\def\rhostar{{\rho^*}}
\def\schema{\mathit{schema}}
\def\var{\mathit{var}}

\newcommand*{\ldb}{\{\mskip-5mu\{}
\newcommand*{\rdb}{\}\mskip-5mu\}}

@@

\title{Parallel Communication Obliviousness: One Round and Beyond}

@@

\begin{abstract}
    This paper studies {\em communication-oblivious} algorithms under the massively parallel computation (MPC) model whose communication patterns follow a distribution that depends only on the number $p$ of machines and the input size $N$ of the underlying problem, regardless of the specific input elements. Our objective is to understand when obliviousness necessitates --- or does not necessitate --- heavier communication compared to the traditional MPC model that does not enforce such a requirement.

    \vgap

    The first part of our investigation focuses on single-round algorithms. We prove that {\em skew-free hashing}, a fundamental problem solvable with load $\tO(N/p)$ with high probability (w.h.p.) under the traditional model, demands a load of nearly $\Omega(N)$ under communication obliviousness. Intriguingly, we show that hashing can still be applied in an oblivious manner to process any natural join in one round w.h.p.\ with a load complexity matching that of the best traditional MPC algorithm. The second part of our investigation studies {\em compilation methods} that convert a traditional MPC algorithm $\A$ into a communication-oblivious counterpart. Given an $\A$ that operates within $\ell = \poly(p)$ rounds with a load at most $L = \Omega(p \log p)$, we can produce a communication-oblivious version running in $2\ell$ rounds with a load at most $(1 + \delta) L$, where $\delta > 0$ can be an arbitrarily small constant. Additionally, we establish hardness results indicating that the theoretical guarantees of our compilation can no longer be significantly improved.
\end{abstract}

@@

In the digital era, cloud computing has emerged as a pivotal part of modern business infrastructure, reshaping the way organizations store, access, and manage data. A growing number of sectors --- ranging from healthcare and finance to academia and entertainment --- are harnessing the power of cloud services to enhance operational efficiency, effectuate cost savings, and fuel innovation. However, the migration of data from local systems to remote servers operated by third-party service providers has ushered in new privacy concerns. The crux of these concerns is the fact that sensitive data, once stored on these servers, is no longer fully under the user's control. Instead, it falls within the stewardship of service providers, thereby raising critical questions about data privacy.

\vgap

Two primary topics come to the forefront in the discourse on privacy control in cloud computing:
\myitems{
	\item {{\bf Confidential computing:}} This aims to mitigate two forms of privacy leakage that might occur on an {\em individual} server. First, rogue administrators could exploit their positions to inspect sensitive data within the server. Second, even if encryption hinders direct data accesses, administrators could still infer valuable information from the memory access patterns that the computation generates.

	\item {{\bf Oblivious communication:}} This aims to prevent another two forms of privacy leakage that could transpire during communication across {\em different} servers. First, malicious network managers might scrutinize the data flowing through the gateways. Second, even with encrypted network traffic, these managers could still deduce sensitive information from the statistics extracted from the communication.
}

Robust solutions to confidential computing are already in place, including systems that employ {\em fully homomorphic encryption} (e.g., IBM's HE4Cloud) or hardware-based {\em trusted execution environments} (nowadays offered by the leading cloud-computing platforms). When integrated with {\em oblivious RAM} techniques \cite{akl+23,go96,ln18,ppry18}, these systems safeguard sensitive information during the entire lifecycle of data: whether the data is stored in the disk, transiting through the memory hierarchy, or in use by the CPU.

\vgap

In this work, we will concentrate on oblivious communication, with the aim to understand how much communication is needed to enforce this requirement. To pave the foundation of our discussion, next we will formalize the concept of oblivious communication and the class of algorithms to be studied.

\subsection{Communication-Oblivious MPC} \label{sec:intro-prob}

Let us first clarify some math conventions in this paper. For an integer $x \ge 1$, the notation $[x]$ denotes the set $\{1, 2, ..., x\}$. Given integers $x > 0$ and $y \ge 0$, let $\binom{x}{y}$ be the number of ways of choosing $y$ elements from a set size $x$; specially, if $y > x$, define $\binom{x}{y}$ $= 0$. We use double-curly braces to represent multi-sets, e.g., $\ldb 1, 1, 1, 2, 2, 3\rdb$ is a multi-set with 6 elements. Every $\log(\cdot)$ has base $2$. The notation $\poly(n)$ denotes the class of functions polynomial in $n$.

\extraspacing {\bf The MPC Model.} Our analysis will be under the {\em massively parallel computation} (MPC) model, which has been extensively adopted in the database literature, as will be surveyed in Section~\ref{sec:intro-prev}.

\vgap

In this model, $p$ share-nothing machines are connected via a high-speed network. Initially, the input data is distributed across the $p$ machines. An algorithm then executes in {\em rounds}, each consisting of two phases: in the first phase, every machine performs computation on its local data, while in the second phase, the machines exchange messages via the network. Importantly, if machine $i \in [p]$ plans to send machine $j \in [p]$ a message $M[i,j]$ (in the second phase), it must prepare the message in the first phase. This restriction prevents machine $i$ from, for example, deciding what to send based on the messages received {\em during} the second phase. The {\em load of a round} is the maximum number of words communicated (sent and received combined) by one machine in that round. The algorithm's performance is measured by two metrics: (i) the total number of rounds executed, and (ii) the maximum load of all rounds, which is termed as the algorithm's (overall) {\em load}.

\vgap

Randomization is modeled by introducing a sequence of random bits, which is agreed upon by all machines. Such an agreement can be reached before receiving the input and therefore does not require communication during the algorithm's execution. In the first phase of a round, each machine can guide its local computation according to the random-bit sequence. When a random event is said to occur ``with high probability'' (w.h.p.), we require that the probability must be at least $1-1/p^c$ where $c$ can be an arbitrarily large constant chosen before seeing the input data.

\vgap

The value of $N$ is commonly assumed to be significantly larger than $p$, which typically means $N \ge p^c$ where $c$ is a constant dependent on the problem studied.

\extraspacing {\bf Communication-Oblivious Algorithms.} We will now formally define the notion of communication obliviousness, building upon a similar formulation in \cite{ccls20}.

\vgap

We consider that an input $\I$ to the underlying problem is a set of elements, each represented with a constant number of words. The {\em problem size} $N$ of $\I$ is defined as the set size $|\I|$. Given a specific value of $N \ge p$, let $\II(N)$ be the collection of all possible inputs with the same problem size $N$. Fix an arbitrary input $\I \in \II(N)$. As mentioned, initially the $N$ elements of $\I$ are stored across the $p$ machines. We say that a partition $\pi = (\I_1, \I_2, ..., \I_p)$ of $\I$ --- that is, $\I_1, ..., \I_p$ are mutually disjoint and their union is $\I$ --- is {\em legal} if assigning $\I_i$ to machine $i$ for each $i \in [p]$ is permitted by the underlying problem.

\vgap

Consider an arbitrary legal partition $\pi = (\I_1, \I_2, ..., \I_p)$ of $\I$ $\in$ $\II(N)$. Suppose that, when executed on $\pi$, an MPC algorithm $\A$ finishes in $\ell$ rounds where, in the $r$-th round ($r \in [\ell]$), machine $i \in [p]$ sends message $\mb{M}_r[i,j]$ to machine $j \in [p]$ (for $i = j$, $\mb{M}_r[i,j]$ is an empty message). Define $\mb{L}_r$ as the $p \times p$ matrix where $\mb{L}_r[i, j]$ equals the length of $\mb{M}_r[i,j]$, measured in the number of words. Every message $\mb{M}_r[i,j]$ is encrypted such that an adversary, who observes the network communication, can only use the sequence %\vspace{-3mm}
\myeqn{
    \sigma = (\mb{L}_1, \mb{L}_2, ..., \mb{L}_\ell) \nn
    %\label{eqn:intro:pattern}
}
to infer about $\I$. We refer to $\sigma$ as the {\em communication pattern} of $\A$ when executed on $\pi$. For a randomized algorithm $\A$, the sequence $\sigma$ may be a random variable.

\vgap

The algorithm $\A$ is {\em communication oblivious} under a specific problem size $N$ if its communication pattern $\sigma$ follows exactly the same probabilistic distribution, regardless of the choice of $\I \in \II(N)$ and the legal partition $\pi$ of $\I$. We say that $\A$ is a {\em single-round algorithm} if $\ell = 1$ for any choice of $\I$ and $\pi$; otherwise, $\A$ is a {\em multi-round algorithm}.

\subsection{Our Contributions} \label{sec:intro:ours}

Our primary objective is to discern the inherent connections between the traditional MPC model and its communication-oblivious counterpart. The core is to understand whether obliviousness would necessitate a provably higher load than the traditional, non-oblivious, MPC model. Next, we provide an overview of our findings.

\extraspacing {\bf Single-Round Hashing.} Our first contribution is a hardness result proving that {\em hashing} --- a crucial building brick deployed extensively by MPC algorithms (in both theory and practice) --- is significantly more expensive under communication obliviousness as far as single-round algorithms are concerned.

\vgap

To pinpoint the source of hardness, we introduce a problem named {\em skew-free gathering}. Fix arbitrary integers $N$ and $p$ such that $N$ is a multiple of $p$, and $p \ge t$ where $t$ is a constant integer at least 2. The input collection $\II(N)$ has only a single input $\I$: this is a set of $N$ elements, each of which --- denoted as $e$ --- is associated with an integer $\key(e)$, referred to as the {\em key} of $e$. The set of keys fulfills two conditions:
\myitems{
    \item There is one special key that is possessed by $t \cdot (N/p)$ elements, called the {\em special elements}.

    \item Every other key is possessed by exactly one element.
}
A partition $\pi = (\I_1, ..., \I_p)$ of $\I$ is {\em legal} if (i) $|\I_i| = N/p$ for all $i \in [p]$, and (ii)
the $t \cdot N/p$ special elements are placed on $t$ distinct machines, i.e., there are $t$ different machine ids $j \in [p]$ such that $\I_j$ contains $N/p$ special elements. The goal of an algorithm is to move all the special elements to one (arbitrary) machine.

\vgap

In the conventional MPC model, the problem can be solved trivially with load $O(N/p)$: simply ask the $t$ machines where the special elements are stored initially to send their data to a common machine. In contrast, we prove that if a single-round communication-oblivious algorithm is required to succeed with probability at least $2/3$, its load must be at least $\Omega(N / p^{1/t})$ with a constant probability. As $t$ increases, the load approaches $\Omega(N)$ up to a factor sub-polynomial in $p$. More precisely, there does not exist any constant $\eps > 0$ such that one can design an algorithm promising a load of $O(N/p^\eps)$ with a constant probability for all $t = O(1)$. We also show that the lower bound $\Omega(N / p^{1/t})$ can be matched by a deterministic algorithm.

\vgap

Let us now turn to {\em skew-free hashing}, where each input in $\II(N)$ is a set $\I$ of $N$ elements, each with an integer key (computed by some hash function). It is guaranteed that every possible key is possessed by $O(N/p)$ elements of $\I$ (hence, skew ``free''). Initially, each machine receives at most $\ceil{N/p}$ elements. The goal is to move all the elements of the same key to a common machine. In the traditional MPC model, the problem can be settled in one round using load $\tO(N/p)$ w.h.p.\ \cite{bks14,bks17c}, where $\tO(\cdot)$ hides a $\polylog p$ factor. As skew-free hashing generalizes skew-free gathering\footnote{Recall that skew-free gathering has the additional constraint that all the elements except for the special ones need to have distinct keys.}, our lower bound on the latter indicates that oblivious skew-free hashing demands a load of nearly $\Omega(N)$, thus creating a huge separation from the traditional model.

%It is worth pointing out that $O(N)$ is the worst possible cost under MPC and can be trivially attained --- obliviously --- by having all machines send their elements to the first machine.

\extraspacing {\bf Single-Round Joins.} The hardness of skew-free hashing under communication obliviousness is alarming because the existing one-round join algorithms \cite{assu13,au11,bks13,bks14,bks17c,kbs16} in the (conventional) MPC model depend on a method named\footnote{Also known as the {\em hyper-cube} method.} \texttt{Share} \cite{au11,bks17c}, which uses hashing as a primitive. Because the community has come very close to resolving the one-round communication complexity of (natural) joins \cite{kbs16}, it would be a huge pity if the topic had to be re-opened in the name of obliviousness.

\vgap

Our second contribution is to show that, fortunately, joins' single-round communication complexities are not affected by obliviousness! To explain, let us first formalize the {\em join problem} under the oblivious setting. Fix an arbitrary integer $N \ge 1$ as the problem size.  Let $\att$ be a finite set, where each element is called an {\em attribute}. Consider a non-empty set $\X \subseteq \att$ of attributes. A {\em tuple} over $\X$ is a function $\bm{u}: \X \rightarrow [N]$. For any non-empty subset $\Y \subseteq \X$, we define the {\em projection} of $\bm{u}$ onto $\Y$, denoted as $\bm{u}[\Y]$, as the tuple $\bm{v}$ over $\Y$ that satisfies $\bm{v}(Y) = \bm{u}(Y)$ for every attribute $Y \in \Y$. A {\em relation} $R$ is a set of tuples over the same set $\Z$ of attributes; we refer to $\Z$ as the {\em schema} of $R$ and represent it as $\schema(R)$. A {\em join} is a set $\Q$ of at least two relations such that $\sum_{R \in \Q} |R| = N$ (i.e., $N$ tuples in total in all the input relations). Define $\schema(\Q) = \bigcup_{R \in \Q} \schema(R)$. The result of $\Q$ is a relation over $\schema(\Q)$, formalized as:
\myeqn{
\join(\Q) = \set{\textrm{tuple $\bm{u}$ over $\schema(\Q)$} \mid \forall R \in \Q: \bm{u}[\schema(R)] \in R}. \nn
}
The {\em schema graph} of $\Q$ is the hypergraph $\G = (\V, \E)$ where
\myeqn{
    \V = \schema(\Q) \text{ and }
    \E = \ldb \schema(R) \mid R \in \Q \rdb. \nn
}
Note that $\E$ is a multi-set because some relations in $\Q$ may have the same schema. Focusing on ``data complexity'', we consider only joins whose schema graphs have constant sizes.

%The quantity $\max_{e \in \E} |e|$ is defined as the {\em arity} of $\G$.

\vgap

Given a hypergraph $\G = (\V, \E)$ and an integer $N \ge 1$ that is a multiple of $p$, we define the class of {\em $(\G, N)$-joins} as the set of joins that have schema graph $\G$ and problem size $N$. In the terminology of communication obliviousness, the collection $\II(N)$ comprises all the $(\G, N)$-joins of the same $\G$. Specifically, each input $\I$ of $\II(N)$ is the set of tuples in the relations of a $(\G, N)$-join $\Q$. We define a partition $\pi = (\I_1, ..., \I_p)$ of $\I$ to be {\em legal} if each $\I_i$, $i \in [p]$, consists of $N/p$ tuples. Given $\pi$, an algorithm $\A$ must output each tuple of $\join(\Q)$ on at least one machine.

\vgap

We show that any $(\G, N)$-join $\Q$ can be settled --- obliviously --- in a single round with load $\tO(N/p^{1/\psi})$ w.h.p.\ where $\psi \ge 1$ is the {\em edge quasi-packing number} \cite{kbs16} of $\G$; see Appendix~\ref{app:hypergraph} for the definition of $\psi$. The load complexity asymptotically matches that of the state-of-the-art single-round algorithm \cite{kbs16} in the traditional MPC model. Like \cite{kbs16}, our algorithm assumes that (i) $N \ge p^c$ where $c$ is a constant dependent on $\G$, and (ii) the machines are aware of certain ``heavy'' values that appear frequently in the relations of $\Q$.

\vgap

Our algorithm applies hashing in a more refined manner compared to \cite{kbs16}. The load $\tO(N/p^{1/\psi})$ stands as an intriguing contrast to our $\Omega(N)$ lower bound (up to a factor sub-polynomial in $p$) on skew-free hashing. This serves as encouraging evidence that hashing can still be powerful in designing single-round algorithms under communication obliviousness. For the join problem, our deployment of hashing circumvents the pitfall of skew-free gathering. Specifically, the deployment ensures that, for each tuple $\bm{u}$ in the join result, the $|\Q|$ tuples in $\set{\bm{u}[\schema(R)] \mid R \in \Q}$ can be transferred to a common machine (which can differ for each $\bm{u}$). Subtly, this is different from sending all tuples having an identical value on an attribute to the same machine, as required by skew-free gathering.

\extraspacing {\bf Round-Doubling Compilation.} The above discussion has focused on single-round algorithms. Our next contribution is a {\em round compilation} method for multi-round algorithms that translates a traditional MPC algorithm $\A$ into the communication-oblivious model while preserving the number of rounds of $\A$ by a factor of 2 and its load by an arbitrarily small constant factor $\delta > 0$. Specifically, if $\A$ performs at most $\ell = \poly(p)$ rounds and, w.h.p., incurs a load of at most $L$, our method yields an oblivious algorithm $\A'$ that computes the same information as $\A$ in $2\ell$ rounds with a load $(1+\delta) L$. Our compilation succeeds w.h.p.\ as long as $L = \Omega(p \log p)$, a condition satisfied by nearly all the existing MPC algorithms we are aware of.

\vgap

We also study how much improvement can still be expected over our compilation method. First, it is clear that the blow-up factor of 2 in the round number cannot be improved in general. As discussed before, for skew-free gathering, it is easy to achieve load $O(N/p)$ in one round in the traditional MPC model, but under communication obliviousness, every single-round algorithm must entail a load of nearly $\Omega(N)$ with a constant probability. Thus, the remaining question is whether the condition $L = \Omega(p \log p)$ can be significantly relaxed. Our last contribution is to answer the question in the negative by establishing a hardness result on the following {\em token passing problem}.

\vgap

Fix arbitrary integers $N$ and $p$ such that $p \ge 2$ and $N = p \cdot \ceil{p^{1-\eps}}$, where the constant $\eps$ satisfies $0 < \eps < 1$. The input collection $\II(N)$ has only a single input $\I$, which is a set of $N$ elements among which there are $N / p = \ceil{p^{1-\eps}}$ special elements called {\em tokens}. In addition to $\I$, the problem takes another two parameters: $x$ and $y$, which are distinct integers in $[p]$. A partition $\pi = (\I_1, ..., \I_p)$ of $\I$ is {\em legal} if (i) $|I_i| = N/p$ for all $i \in [p]$, and (ii) the $N/p$ tokens are all placed on machine $x$ (i.e., $I_x$ consists of nothing but tokens). The objective of the problem is to move all the tokens to machine $y$.\footnote{Unlike skew-free gathering where all the special elements can be moved to an arbitrary machine, here all the tokens must be sent to machine $y$.}

\vgap

In the traditional model, the problem can be trivially solved in one round with load $L = O(p^{1-\eps})$: simply ask machine $x$ to send all the tokens to machine $y$. We prove that, if a two-round communication oblivious algorithm is required to succeed with probability at least 2/3, its load must be $\Omega(p^{1-\eps/2})$ with a constant probability. The polynomial gap between $\Omega(p^{1-\eps/2})$ and $O(p^{1-\eps})$ indicates that the condition $L = \Omega(p \log p)$ of our round-doubling compilation  is tight up to a factor sub-polynomial in $p$. In fact, this remains true even if the blow-up factor $1+\delta$ in load preservation is increased to $\polylog p$.

\subsection{Previous Results and Relevance to Ours} \label{sec:intro-prev}

Closely related to our work is a compilation method by Chan et al.\ \cite{ccls20}. Our compilation draws inspiration from theirs but makes several improvements. First, the approach of \cite{ccls20} was designed for the ``SODA MPC'' model \cite{ksv10}, which resembles the model in Section~\ref{sec:intro-prob} but does away the notion of ``load''. In that model, each machine is equipped with $s = \Omega(N/p)$ words of memory, and the amount of communication is not a main concern as long as each machine receives no more than $s$ words in each round from all other machines combined. The analysis of \cite{ccls20} was carried out under the condition of $s = N^\eps$ for some constant $\eps > 0$. When translated into our scenario, the analysis fails to capture the regime where $L \ll s$. Second, our compilation is substantially simpler, restores the clarity for the approach underneath \cite{ccls20} in the scenario where $L = \Omega(p \log p)$, and explicitly determines the blow-up factors $1+\delta$ and 2 (for preserving the load and number of rounds, respectively)\footnote{In \cite{ccls20}, these factors were hidden in big-$O$.}. Finally, our discussion (through the token passing problem) on the condition that $L$ needs to satisfy to enable round-doubling load-preserving compilation is new.

\vgap

It should be further noted that the study in \cite{ccls20} does not address one-round MPC algorithms. Through our round compilation method, one can see that a primary distinction between the conventional MPC model and its oblivious counterpart lies in the realm of one-round algorithms. Exploring fundamental problems that can characterize this difference represents an interesting direction. Our work can be seen as a step in that direction.

\vgap

As mentioned, the existing single-round algorithms for join processing \cite{assu13,au11,bks13,bks14,bks17c,kbs16} rely crucially on the \texttt{Share} method \cite{au11,bks17c}. Given a join $\Q$, \texttt{Share} assigns a hash function $h_X(\cdot)$ to each attribute $X \in \schema(\Q)$. For each tuple $\bm{u}$ in a relation $R \in \Q$, the machine initially storing $\bm{u}$ transmits it to a set of machines determined by the hash values in $\set{h_X(\bm{u}(X)) \mid X \in \schema(R)}$. Beame et al.\ \cite{bks17c} presented a sharp concentration bound on the load of \texttt{Share} when certain skew-free requirements are satisfied. Their bound is useful to our analysis and will be reviewed in Section~\ref{sec:join}. Koutris et al.\ \cite{kbs16} proved that any one-round MPC algorithm (oblivious or not) solving the join problem must incur a load of $\Omega(N/p^{1/\psi})$ with a constant probability, if each machine is aware of only the tuples in its local storage initially.

\vgap

There is an extensive body of literature on multi-round MPC algorithms; see \cite{ajr+17,aksu15,akz19,hyt19,kst22,hcww21,hk20,h21,hy19,qt21,t18b} and the references therein. None of those algorithms was designed to achieve communication obliviousness, but all of them satisfy the condition $L = \Omega(p \log p)$ and, therefore, can be made oblivious by our compilation method. It is worth noting that our compilation also works on the one-round algorithms in \cite{assu13,au11,bks13,bks14,bks17c,kbs16}, but the algorithms after compilation will require two rounds.

\section{Hardness of Skew-Free Gathering} \label{sec:gather}

This section will discuss the skew-free gathering problem defined in Section~\ref{sec:intro-prob} and serves as a proof of the paper's first main result:

\begin{theorem} \label{thm:gathering}
    For the skew-free gathering problem parameterized by $t$, any communication-oblivious one-round algorithm succeeding with probability at least $2/3$ must entail a load of $\Omega(N/p^{1/t})$ with at least a constant probability.
    %Furthermore, this is tight: there is a communication oblivious algorithm that always succeeds and guarantees a load of $O(N/p^{1/t})$ deterministically.
\end{theorem}

In Appendix~\ref{app:sf-gathering-alg}, we show that the above lower bound $\Omega(N/p^{1/t})$ can be matched deterministically.

\vgap

To prove Theorem~\ref{thm:gathering}, let us start by identifying a set $\Pi$ of $\binom{p}{t}$ legal partitions of the (sole) input $\I \in \II(N)$ as follows. First, choose a set $Z$ of distinct integers $z_1, z_2, ..., z_t$ in $[p]$; we will refer to $Z$ as the {\em seed machine set}. Second, place $N/p$ special elements on machine $z_j$ for each $j \in [t]$, and place $N/p$ innocuous elements on each machine $i \in [p] \setminus Z$, ensuring that every  element is on exactly one machine. This defines a legal partition in $\Pi$. As there are $\binom{p}{t}$ ways to choose $Z$, the size of $\Pi$ is $\binom{p}{t}$. We require a one-round algorithm $\A$ to be oblivious only on $\Pi$, namely, its communication pattern follows the same probabilistic distribution when it is executed on $\pi \in \Pi$. We will argue that the load of $\A$ must still be $\Omega(N/p^{1/t})$ with at least a constant probability if $\A$ succeeds on every $\pi \in \Pi$ with probability at least $2/3$. This is sufficient for validating the lower bound in Theorem~\ref{thm:gathering}.

\extraspacing {\bf Conditional Expected Load.} As $\A$ executes in one-round, its communication pattern can be fully characterized by a single $p \times p$ matrix $\mb{L}$ where $\mb{L}[i,j]$ ($i, j \in [p]$) is the length of the message $M[i,j]$ that machine $i$ sends to machine $j$. The load of $\A$ can now be calculated as
\myeqn{
    \load(\mb{L}) &=&
    \max_{i=1}^p
    \Big(
    \sum_{j=1}^p \mb{L}[i, j] + \mb{L}[j, i]
    \Big).
    \label{eqn:gathering:load-matrix}
}
When $\A$ is randomized, the $M[i,j]$ of all $i, j \in [p]$ are random variables, and hence so are $\mb{L}$ and
$\load(\mb{L})$. Our objective is to prove that $\Pr[\load(\mb{L}) = \Omega(N/p^{1/t})]$ is at least a constant.

\vgap

By communication obliviousness, the distribution of $\mb{L}$ is the same for all the legal partitions in $\Pi$. Define $\L$ as the set of $p \times p$ matrices $\mb{\Lambda}$ satisfying
\myitems{
    \item $\Pr[\mb{L} = \mb{\Lambda}] > 0$, namely, $\A$ exhibits the patten $\mb{L} = \mb{\Lambda}$ with a non-zero probability;
    \item $\max_{i=1}^p \sum_{j=1}^p \mb{\Lambda}[i, j] \le N/p^{1/t}$;
    \item $\max_{i=1}^p \sum_{j=1}^p \mb{\Lambda}[j, i] \le N/p^{1/t}$.
}
Note that when $\mb{L} = \mb{\Lambda}$, {\em every} machine sends/receives at most $N/p^{1/t}$ words. The subsequent discussion will assume
\myeqn{
    \Pr[\mb{L} \in \L] \ge 1/2. \label{eqn:gathering:small-load-1/2}
}
Note that whenever $\mb{L} \notin \L$, some machine communicates at least $N/p^{1/t}$ words and hence $\load(\mb{L}) \ge N/p^{1/t}$. Thus, $\Pr[\mb{L} \notin \L] \ge 1/2$ immediately gives $\Pr[\load(\mb{L}) \ge N/p^{1/t}] \ge 1/2$, thereby verifying the claim in Theorem~\ref{thm:gathering}.

\vgap

We will aim to validate the following lower bound on the expectation of $\load(\mb{L})$ subject to the condition $\mb{L} \in \L$:
\myeqn{
    \expt[\load(\mb{L}) \mid \mb{L} \in \L] &\ge& c \cdot N/p^{1/t}
    \label{eqn:gathering:goal}
}
for some constant $0 < c < 1$. The above implies\footnote{For every $\mb{L} \in \L$, it holds that $\load(\mb{L}) \le 2 N/p^{1/t}$. If \eqref{eqn:gathering:goal-implication} incorrect, then $\expt[\load(\mb{L}) \mid \mb{L} \in \L] <  \fr{2N}{p^{1/t}} \cdot \fr{c}{4} + \fr{c}{2} \fr{N}{p^{1/t}} (1 - \fr{c}{4}) < c N / p^{1/t}$, contradicting \eqref{eqn:gathering:goal}.}
\myeqn{
    \Pr[\load(\mb{L}) \ge (c/2) \cdot N/p^{1/t} \mid \mb{L} \in \L] \ge c/4.
    \label{eqn:gathering:goal-implication}
}
Combining \eqref{eqn:gathering:small-load-1/2} and \eqref{eqn:gathering:goal-implication} yields $\Pr[\load(\mb{L}) \ge \fr{c}{2} N/p^{1/t}] \ge \fr{c}{4} \cdot \fr{1}{2} = c/8$, thus establishing Theorem~\ref{thm:gathering}.

\extraspacing {\bf Heavy Senders.} Suppose that the algorithm $\A$ exhibits a pattern $\mb{L} = \mb{\Lambda} \in \L$. Given a machine $i \in [p]$, we call machine $j \in [p] \setminus \set{i}$ a {\em heavy sender} for machine $i$ under $\mb{\Lambda}$ if $\mb{\Lambda}[j,i] \ge N/p$, namely, machine $j$ sends at least $N/p$ words to machine $i$. Define
\myeqn{
    H(\mb{\Lambda}, i) = \text{number of heavy senders for machine $i$ under $\mb{\Lambda}$}. \label{eqn:gathering:H}
}
It is easy to see that
\myeqn{
    H(\mb{\Lambda}, i) \le p^{1-1/t}
    \label{eqn:gathering:bound-H}
}
because otherwise machine $i$ would receive more than $(N/p) \cdot p^{1-1/t} = N/p^{1/t}$ words, contradicting $\mb{\Lambda} \in \L$.

\vgap

Our analysis will revolve around heavy senders. We will argue that at least one machine $i \in [p]$ satisfies
\myeqn{
    \sum_{\mb{\Lambda} \in \L} \Pr[\mb{L} = \mb{\Lambda}] \cdot H(\mb{\Lambda}, i)
    &=&
    \Omega(p^{1-1/t}).
    \label{eqn:gathering:goal2}
}
As $\load(\mb{\Lambda}) \ge (N/p) \cdot H(\mb{\Lambda}, i)$, the above leads to
\myeqn{
    \sum_{\mb{\Lambda} \in \L} \Pr[\mb{L} = \mb{\Lambda}] \cdot \load(\mb{\Lambda})
    =
    \Omega\Big(\fr{N}{p} \cdot p^{1-1/t} \Big)
    =
    \Omega(N /p^{1/t}).
    \label{eqn:gathering:goal2-implication}
}
This will prove \eqref{eqn:gathering:goal} because
\myeqn{
    \expt[\load(\mb{L}) \mid \mb{L} \in \L]
    &=&
    \fr{1}{\Pr[\mb{L} \in \L]} \sum_{\mb{\Lambda} \in \L} \Pr[\mb{L} = \mb{\Lambda}] \cdot \load(\mb{\Lambda})
    \nn
}
which is at least \eqref{eqn:gathering:goal2-implication}.

%as long as $H(\mb{\Lambda}, i) \ge 1$

\extraspacing {\bf Two Types of Legal Partitions.} The rest of the section will focus on proving the correctness of \eqref{eqn:gathering:goal2}. Earlier we have identified a set $\Pi$ of $\binom{p}{t}$ legal partitions, each characterized by a seed machine set $Z = \set{z_1, ..., z_t}$; for convenience, we will use $Z$ to denote the corresponding legal partition when no ambiguity can arise. Now, fix any legal partition $Z$. The algorithm $\A$ succeeds on $Z$ if and only if at least one of the following occurs:
\myitems{
    \item machines $z_1, z_2, ..., z_t$ all send their special elements to a machine $k \in [p] \setminus Z$;
    \item there is some $j \in [t]$ such that machine $z_j$ receives all the special elements from every other machine in $Z$.
}
Echoing the above, given any $Z$ and any matrix $\mb{\Lambda} \in \L$, we define
\myitems{
    \item $X(Z, \mb{\Lambda})$ as an indicator variable, which equals 1 if there exists some $k \in [p] \setminus Z$ such that $\mb{\Lambda}[z_j, k] \ge N/p$ for every $j \in [t]$, or 0 otherwise;

    \item $Y(Z, \mb{\Lambda})$ as an indicator variable, which equals 1 if there exists some $j \in [t]$ such that $\mb{\Lambda}[z_k, z_j] \ge N/p$ for every $k \in [t] \setminus \set{j}$, or 0 otherwise.
}

%

When exhibiting the pattern $\mb{L} = \bm{\Lambda}$, the algorithm $\A$ succeeds on $Z$ only if
\myeqn{
    X(Z, \mb{\Lambda}) + Y(Z, \mb{\Lambda}) \ge 1. \nn
}
We declare:
\myeqn{
    \sum_{\mb{\Lambda} \in \L} \Pr[\mb{L} = \mb{\Lambda}] \cdot
    (X(Z, \mb{\Lambda}) + Y(Z, \mb{\Lambda}))
    &\ge&
    1/6.
    \label{eqn:gathering:1/6}
}
To see why, notice that if the above did not hold, then $\A$ would have probability less than $1/6$ succeeding on $Z$ when $\mb{L} \in \L$. Even if $\A$ always succeeds on $Z$ when $\mb{L} \notin \L$, the overall success probability on $Z$ would still be less than $1/6 + \Pr[\mb{L} \notin \L]$, which is less than 2/3 because of \eqref{eqn:gathering:small-load-1/2}. This violates the requirement that $\A$ must succeed with probability at least $2/3$ on $Z$.

\vgap

We will refer to a legal partition $Z$ as
\myitems{
    \item an {\em X-partition} if $\sum_{\mb{\Lambda} \in \L} \Pr[\mb{L} = \mb{\Lambda}] \cdot
    X(Z, \mb{\Lambda}) \ge 1/12$, or
    \item an {\em Y-partition} if $\sum_{\mb{\Lambda} \in \L} \Pr[\mb{L} = \mb{\Lambda}] \cdot
    Y(Z, \mb{\Lambda}) \ge 1/12$.
}
Note that $Z$ must be classified as at least one of the two types because of \eqref{eqn:gathering:1/6}. Thus, if $\XX$ (resp., $\YY$) represents the set of all X- (resp., Y-) partitions, we have $|\XX| + |\YY| \ge |\Pi| = \binom{p}{t}$. The next lemma relates each type of partitions to heavy-sender machines.

\begin{lemma} \label{lmm:gathering:total-X-Y}
    For any $\mb{\Lambda} \in \L$, it holds that
    \myeqn{
        \sum_{Z \in \XX} X(Z, \mb{\Lambda}) &\le& \sum_{i \in [p]} \binom{H(\mb{\Lambda}, i)}{t} \label{eqn:gathering:total-X-Y:help1} \\
        \sum_{Z \in \YY} Y(Z, \mb{\Lambda}) &\le& \sum_{i \in [p]} \binom{H(\mb{\Lambda}, i)}{t-1}. \label{eqn:gathering:total-X-Y:help2}
    }
\end{lemma}

The reader is reminded that $\binom{x}{y}$ equals 0 if $x < y$. We will prove the first inequality here and the second one in Appendix~\ref{app:proof:thm:gathering}.

\begin{proof}[Proof of \eqref{eqn:gathering:total-X-Y:help1}]
    Our proof adopts a counting argument. Initially, create an empty set $S_i$ for each $i \in [p]$. Process each $Z = \set{z_1, ..., z_t} \in \XX$ as follows. If $X(Z, \mb{\Lambda}) = 0$, do nothing. Otherwise, under the communication pattern $\mb{L} = \bm{\Lambda}$, there is at least one machine $k \in [p] \setminus Z$ that receives $N/p$ special elements from machine $z_j$ for all $j \in [t]$. Thus, machines $z_1, z_2, ..., z_t$ are heavy senders for machine $k$ under $\mb{\Lambda}$. Hence, $Z$ represents a possible way to choose $t$ machines from the $H(\mb{\Lambda}, k)$ heavy senders for machine $k$ under $\mb{\Lambda}$. Accordingly, we add $Z$ to $S_k$. After all the sets $Z \in \XX$ have been processed, the left hand side of \eqref{eqn:gathering:total-X-Y:help1} is bounded by $\sum_{i \in [p]} |S_i|$. On the other hand, for each $i \in [p]$, the size $|S_i|$ cannot exceed the total number of ways of choosing $t$ machines from the $H(\mb{\Lambda}, i)$ heavy senders for machine $i$ under $\mb{\Lambda}$. Inequality \eqref{eqn:gathering:total-X-Y:help1} now follows.
\end{proof}

>>>>

@@

\title{Optimal Multiway Spatial Joins}

@@

\begin{abstract}
    Later.
\end{abstract}

@@
